Tutorial: Fitting log-Gaussian mixed-effects mixture models for keystroke data

### 2024-04-28 report/sections/introduction.Rmd
manuscript update

Translating ideas into language involves a cascade of processes starting with a communicative intention, over the generation of an message, deciding with which part of the message to start the utterance, retrieving appropriate lexical materials and ordering them in an appropriate syntactic order before retrieving orthographic and / or phonological codes that are finally submitted to the motor programs that allow us to articulate language in speech, keyboard tying, handwriting, and signing [@bock2014syntactically; @wheeldon_konopka_2023; @van1991handwriting; @olive2014toward].

Writing is arguably more demanding than spoken language, because, among other reasons, writing requires spelling and we typically do not move our fingers or our pen as quickly as we articulate language sounds in speech. Consequently in writing more than in speech, we have to mentally buffer our linguistic plan before we can express it in language. Written communication is also less time constrained: speaking underlies fluency requirements and therefore has stronger pressure to plan information ahead of speaking [@roeser2018advance; @torrance2012written]. This is because in spoken production there is a pressure for fluency whereas hesitations in written output have no communicational implications for the reader [e.g. @Clark2002]. Indeed writers can pause when -- even in the middle of a word -- and for how long they want (within reason) without compromising communication. Written text production -- for example argumentative texts -- is therefore a combination for relatively fluent production bursts followed by pauses [@chenoweth2001fluency].

The study reported in this paper presents a comparison of two theories of how cognitive planning processes are coordinated in writing using implementations in statistical models. On the basis of those statistical models applied to 6 datasets, we revisit the often reported finding that edges of larger linguistic units are associated with longer pauses [@conijn2019understanding; @matsuhashi1981pausing; @flower1981pregnant; @wen06]. For example, sentence boundaries are generally thought of as involving longer pauses, compared to word boundaries, because participants engage in higher-level planning [@baaijen2012keystroke; @roeser2018advance; @medimorec2017disfluency; @medimorec2017pauses]. The same is generally observed for spoken utterances [@lee13; @meyer1996; @wheeldon2013] with some exceptions [@gri01]. We first discuss pause patterns in text production and introduce two possible cognitive frameworks in which these can be interpreted. Pause-patterns at linguistic edges are important, as they are symptomatic for the two cognitive theories of writing we discuss in the following.

Written text production is the result of a cascade of processes that starts with an communicative idea on the top forming a semantic representation, which is then fed into a grammatical encoder that is responsible for the retrieval of lexical names and, depending on theoretical viewpoint [@bock2014syntactically; @wheeldon_konopka_2023], the generation of syntactic representations, and finally flow from orthographic retrieval into motor plans for executing finger movements [@van1991handwriting; @olive2014toward]. The intervals between adjacent keystrokes during typing vary as a function of various factors. Explaining the factors that influence the duration of inter-keystroke intervals (IKI) requires a theory of how the various processes that transform intent into keypresses are structured. Processing may be serial. This means that the processing cascade operates on one unit at a time. Processing of the next unit can only start when the previous unit was fully processed from communicative intention to completion of the written output. Therefore longer IKIs represent periods when the writer is planning what to write next. The subsequent burst of fluent output is then the result this planning [@hayes2012evidence; @matsuhashi1981pausing; @alves2015progress; @schilperoord2002cognitive; @kaufer1986composing]. This is particularly reflected in the model proposed by @flower1980ynamics that characterises the writing process as a sequence of generating ideas, planning the text, translating the ideas into language, executing hand-writing or typing, and revising the texts [also @chenoweth2001fluency; @chenoweth2003inner].

This serial account is consistent with the finding that average IKI duration immediately before sentence-initial keystrokes are longer than before words and these are, in turn, longer than between mid-word key presses [@torrance2016adolescent; @wen02; @mohsen2021second]. However this account is not consistent with what we know about writing. Consider these two observations:

  First, utterances are not fully formed at production onset. Instead, syntax and lexical content and even details of the message itself are planned as the emergent result of in real-time process. This is well known for spoken production [e.g. @bock2014syntactically]. For writing, there is evidence that the time to keystroke / pen onset increases when a sentence starts with a more complex sentence-initial phrases [@not10; dam09; @roeser2018advance]. For example, in @roeser2018advance participants described moving arrays of images in simple utterances such as *A and the B moved above the C* that either started with a conjoined noun phrase (i.e. *A and the B moved*) or a simple noun (i.e., *A moved above the B and the C*). Importantly while the overall complexity of the utterance (in terms of length, number of phrases and noun) was held constant, sentences that started with a conjoined noun phrase increased the time to typing onset. This is not what one would expect if writers plan sentences (or even phrases) in full before production onset. In fact, eye movement data reported in @torrance2012written and @roeser2018advance demonstrate that writers do not plan the lexical form of the utterance beyond the first noun before production onset [similar to speech @gri01; @griffin2003reversed], although there was evidence that lexical pre-planning of the second noun (i.e. *the B*) was more likely for conjoined noun phrases.

Second, even though average sentence-initial IKIs tends to be longer, they still tend to be very rapid. For example, @medimorec2017pauses found that, in undergraduate students writing on familiar topics 71% of sentence-initial IKIs were less than 1 sec in duration. Also @ronneberg2022process reported that sentence-initial hesitations are rare, with over 50% of sentences preceded by very short pauses (around 430 msecs) and a mean of 1.2 seconds for the remainder of pauses. For comparison, this is less than mean written naming response latency for single objects [@torrance2018timed] and short sentences when describing arrays of images [@roeser2018advance] in a similar population. Despite the fact that writers can in principle pause when they want to, written composition is often remarkably fluent. The finding that sentence-initial IKIs are longer before words, maps onto a theory that assumes that planning a sentence requires the writer to pause and think before pressing the sentence-initial key.

If utterances are not planned in full prior to writing onset, how is it possible that at least for reasonably competent writers producing composition often occurs remarkably fluently, with very few hesitations. The two examples above and similar findings point towards much of the mental activity associated with composition, including the relatively complex processing required to plan sentences, occurring as a result of a cascade of processes that occur partly in parallel and largely without executive control. Consistent with general trends in language processing theory [@dell2007case; @bock2014syntactically; @chang2006], several researchers have argued that the processes associated with written production run at least in part in parallel [@bonin2012evidence; @crump2010hierarchical; @olive2014toward; @van1991handwriting; @roux2013interaction]. @van1991handwriting, for example, argued for a cascade of modular processes, each responsible for a specific transformation (semantic, syntactic, and so forth), with processing occurring as soon as information from the immediately upstream process becomes available [@christiansen2016now]. Buffers provide transient storage at each processing level to accommodate unsynchronised output rates: when lower level processes lag behind higher level processes, buffers allow our fingers to catch up with thought and language related processes. In other words, when writers move from one sentence to another without hesitation this is due to planning the next sentence, to some extent, while completing output of the previous sentence.

This cascaded account of the composition process gives a rather different understanding of inter-keystroke intervals. IKIs result from one of two data-generating processes. When upstream processes output at a rate equal to or faster than can be used for finger movement planning then IKIs are determined just by time needed for executing finger movements (i.e. just by processing below the last buffer in the processing cascade). However if one or more upstream processes provide output more slowly then IKIs become dependent directly on upstream processing times and not on finger movement. IKIs therefore form two distributions, one associated with rapid, fluent output, and another that forms as a result of delays caused by some combination of processing at the semantic, syntactic, lexical, or orthographic levels [@roeser2021modelling].

For example, @roeser2021modelling demonstrated -- using a similar approach as used in the present paper -- how this mixture of distributions that results from the same underlying data generating process can be implemented using Bayesian statistical mixture modelling. In particular they analysed a large sample of copy-task data using commonly used statistical models and compared their out-of-samples predictive performance to a mixture model approach. Not only did they find substantially stronger performance for their mixture model approach but they also demonstrated that mixture models allowed them to capture the process that generates IKIs in a copy-task context in three parameters: (1) the average speed of fluent key transitions, (2) the slowdown for hesitant key presses, and (3) the probability of hesitant as opposed to fluent transitions. As argued by the authors, copy-typing underlies a similar cascade of processes as free text composition. Instead of context generation and grammatical encoding, copy-typing involves a mixture of visual encoding of the target string and mental buffering prior to the activation of motor codes. For similar approaches see @almond2012preliminary, @baaijen2012keystroke, @chenu2014interword, @guo2018modeling, @hall2022constructing, @li2021identifying, and @waes2019.

The cascading account provides an explanation for why writers sometimes pause before onsetting a sentence (or word) but often are not. This is not because in some cases sentences are not planned, or planning was postponed until after sentence onset, but that this planning was completed in parallel with previous output.

Hesitations in writing have potential disadvantages. Writers can, as we have noted, pause at any point without this affecting the eventual communicative effect of their text. However language processing is subject to what @christiansen2016now describe as a fundamental "now-or-never" bottleneck. Buffering is transient and as a consequence of written (and spoken) production are just-in-time systems: Production must flow down the cascade of processes from message to finger movements without significant interruptions. If there is delay -- if, for example, a writer struggles to retrieve the right word or its spelling -- there is a risk that output from upstream processes, and particularly the chain of ideas that the writer wanted to communicate, will be lost. @ronneberg2022process coined this possibility the "process-disruption hypothesis".

In the study that is presented in this paper we test using Bayesian statistical hierarchical models of keystroke intervals -- similar to @roeser2021modelling -- whether the cascading view of writing generalises to contexts in which participants compose texts. On the basis of these statistical models of text composition, we appraise to what extent linguistic edges (sentence and word boundaries) are associated with different pausing behaviours as frequently suggested in the literature [@ailhaud2018variations; @ailhaud2016developmental; @conijn2019understanding; @chukharev2019combined; @torrance2016adolescent; @wen02; @mohsen2021second].

Data form spontaneous text production [see @gernsbacher1995coherence] are an ideal test case for our modelling approach for the following three reasons: First, text production such as essay, argumentative, and narrative writing in response to a writing brief or topic statement is a natural environment for text production. Therefore our models have real-world relevance for educators, for example, to evaluate to what extent a student is using a desirable composition strategy [@dux2022automating; @dux2021effect; @vandermeulen2023writing].

Second, since the emergence of software such as InputLog [@leijten2013keystroke] recording keystroke data during free text composition has become widely used among writing researchers and led to a rapidly expanding timecourse literature with a focus on the orchestration of subprocesses; with a recent special issue in Reading and Writing [@conijn2021timecourse]. Social Science Citation Index reports 40 journal papers that describe research exploring composition processes using keystroke logging methods in 2022, compared to 26 in 2017 to 2019, and 7 in 2014 to 2016. These data are therefore important for contemporary theories of writing and freely available for modelling and machine learning [@conijn2019understanding; @conijn2020keys].

Third, data from spontaneous text production allow us to address questions about sentence production in context that would not be possible with data from, for example, picture description experiments [@not10; @dam09; @roeser2018advance] or copy tasks [@roeser2021modelling; @waes2019; @van2019multilingual]. In text composition, content generation is not artificially constrained to individual sentence units (or less) by one or more images -- as in sentence-elicitation studies -- and linguistic form is not constraint to a sequence of words -- as in copy tasks. Instead text composition provides data from both inter and intra-sentence transitions. Also, pauses do not reflect visual encoding of the stimulus (although possibly looking back during writing which we will return to in the discussion).

The hypotheses are as follows: if the preparation of the upcoming production unit happens entirely at the corresponding linguistic edge as predicted by the serial account, and not in parallel to production, key-transition intervals can statistically be modelled as a function of the associated transition location. In other words, we expect IKIs to be proportionally longer at before-sentence locations compared to before-word location compared to within-word locations. This is because transitions before larger linguistic units are associated with processing that involves higher levels of representation. The cascading view, in contrast, assumes that although IKIs at larger linguistic edges might be longer, it is in principle possible that writers do not pause but plan in parallel to writing. In other words, statistical models of the pause-and-burst view of written production do not capture that planning can operate, to some extent, in parallel to the output of the written production. Therefore, a statistical model that captures writing as a cascade of processes must account for the possibility that hesitations occure probabilistic across the entire text although some linguistic locations such as larger linguistic edges are associated with a higher pausing probability. We hypothesize that the statistical model of the parallel cascading view provides a better out-of-sample generalization than the implementation of the serial model of writing in the context of unconstrained text composition.

>>>>

Translating ideas into language involves a sequence of processes starting with communicative intent, proceeding through generation of a message, retrieving appropriate lexical items, deciding with which part of the message to start the utterance which, depending on theoretical viewpoint [@bock2014syntactically; @wheeldon_konopka_2023], is determined by syntactic requirements of our language, then retrieving orthographic and / or phonological codes, and that are finally submitted to the motor programs that output speech, key or pen-strokes or signs [@wheeldon_konopka_2023; @van1991handwriting; @olive2014toward].

The coordination of these processes in writing is arguably more demanding than in spoken language. This is because writing requires spelling retrieval which may or may not be mediated by phonology, and the motor planning and execution required to press keys or move a pen are later-developing and more effortful than those required to articulate language sounds in speech leading to a reduced output speed compared to speech [see for a recent review @kandel2023written]. Consequently when writing, more than in speech, we have to mentally buffer linguistic plans before we can express them in language. Written communication is also less time constrained: speaking underlies fluency requirements, because hesitating during speech has a communicative effect [e.g. @Clark2002]. There is therefore stronger pressure to plan ahead in speech than in writing [@roeser2018advance; @torrance2012written]. Indeed writers can pause when they wish -- even in the middle of a word -- and for how long they want (within reason) without compromising communication. Written text production -- for example writing argumentative texts -- is therefore a combination for relatively fluent production bursts followed by pauses [@chenoweth2001fluency].

The study reported in this paper compares statistical models that instantiate two different theories of how cognitive planning processes are coordinated in writing. Through the analysis of 6 different datasets from writers of different ages and competencies composing multi-sentence texts, we compare these statistical models and therefore the cognitive model of how keystrokes in writing are being generated they are mapping on. Also, we evaluate competing predictions of these models by revisiting the often-reported finding that edges of larger linguistic units are associated with longer pauses [@conijn2019understanding; @matsuhashi1981pausing; @flower1981pregnant; @wen06]. In what follows we first summarise previous research that has explored pause patterns in spontaneous text production. We then introduce two possible cognitive frameworks in which these can be interpreted. Finally we argue that analysis of pause-patterns at linguistic edges are important to permit differentiation between these two theories.

We focus on data form spontaneous text production [see @gernsbacher1995coherence] for the following three reasons: First, text production such as essay, argumentative, and narrative writing in response to a writing brief or topic statement is a natural environment for text production and have, therefore, real-world relevance [@dux2022automating; @dux2021effect; @vandermeulen2023writing]. Second, since the emergence of keystroke-logging software such as InputLog [@leijten2013keystroke] recording keystroke data during free text composition has become widely used among writing researchers and led to a rapidly expanding timecourse literature with a focus on the orchestration of subprocesses; for a recent special issue see @conijn2021timecourse.^[Social Science Citation Index reports 40 journal papers that describe research exploring composition processes using keystroke logging methods in 2022, compared to 26 in 2017 to 2019, and 7 in 2014 to 2016.] These data are therefore important for contemporary theories of writing and freely available for statistical modelling and machine learning [@conijn2019understanding; @conijn2020keys]. Third, spontaneous text production -- unlike picture description experiments [@not10; @dam09; @meulemans2021; @roeser2018advance] or copy tasks [@roeser2021modelling; @waes2019; @van2019multilingual] -- content generation is not artificially constrained to individual sentence units (or less) by one or more images and linguistic form is not constrained to a sequence of words, respectively. Instead text composition provides data from both inter and intra-sentence transitions.^[Also, pauses in text production do not reflect visual encoding of the stimulus (although possibly reflect looking back during writing which we will return to in the discussion).]

Time intervals between adjacent keystrokes during typing vary as a function of various factors. Explaining the factors that influence the duration of inter-keystroke intervals (IKIs) requires a theory of how the various processes that transform intent into keypresses are structured. Processing -- planning what to say and how to say it -- may be serial. This means that the processing sequence operates one unit at a time. Processing of the next language unit can only start when the previous unit was fully processed from communicative intention to completion of the written output. Therefore longer IKIs represent periods when the writer is planning what to write next. The subsequent burst of fluent output is then the result this planning [@hayes2012evidence; @matsuhashi1981pausing; @alves2015progress; @schilperoord2002cognitive; @kaufer1986composing]. This is particularly reflected in the model proposed by @flower1980ynamics that characterises the writing process as a sequence of generating ideas, planning the text, translating the ideas into language, executing hand-writing or typing, and revising the texts [also @chenoweth2001fluency; @chenoweth2003inner].

This serial account is consistent with the frequent finding that average IKI duration immediately before larger linguistic units compared to smaller linguistic units. For example, in spontaneous text the mean pause duration before a writer starts a sentence has typically been reported to be longer than before a mid-sentence word and these are, in turn, longer than between mid-word key presses [@torrance2016adolescent; @wen02; @mohsen2021second; @miller2006pausing; @xu2017analyzing; @miller2008psycholinguistic;@deane2015exploring]. This is understood as being due to time needed to plan message and syntax for the next sentence, which is added to the time required to prepare the upcoming word, and the motor planning required to produce the first keystroke of that word [@baaijen2012keystroke; @roeser2018advance; @medimorec2017disfluency; @medimorec2017pauses]. However this account is not consistent with what we know about writing. Consider these two observations:

First, writers and speakers can in principle plan utterances in full for one or multiple sentences before beginning to output words. However this is not what people typically do: utterances are not fully formed at production onset. Instead, syntax and lexical content and even details of the message itself are planned as the emergent result of a real-time process. For speech it is well known that the pre-planning duration depends -- among other factors -- on the need to avoid intra-sentence hesitation that are needed to plan upcoming words in parallel to production [e.g. @levelt2000word; @wheeldon_konopka_2023]. For writing, there is evidence that the time to keystroke / pen onset increases when a sentence starts with a more complex sentence-initial phrases similar as in speech [@not10; dam09; @roeser2018advance]. For example, in @roeser2018advance participants described moving arrays of images in simple utterances such as *A and the B moved above the C* that either started with a conjoined noun phrase (i.e. *A and the B moved*) or a simple noun (i.e., *A moved above the B and the C*). Importantly while the overall complexity of the utterance (in terms of length, number of phrases and noun) was held constant, sentences that started with a conjoined noun phrase increased the time to typing onset (and speech onset). This is not what one would expect if writers plan sentences in full before production onset. In fact, eye-movement data reported in @torrance2012written and @roeser2018advance demonstrate that writers do not plan the lexical form of the utterance beyond the first noun before production onset [similar to speech @gri01; @griffin2003reversed], although there was evidence that lexical pre-planning of the second noun (i.e. *the B*) was more likely for conjoined noun phrases, i.e. when the second noun was part of the same phrase as the first noun.

Second, even though the average sentence-initial IKI tends to be longer, they still tend to be very rapid. For example, @medimorec2017pauses found that, in undergraduate students’ writing on familiar topics 71% of sentence-initial IKIs were less than 1 seconds in duration. Also @ronneberg2022process reported that sentence-initial hesitations are rare, with over 50% of sentences preceded by very short pauses (around 430 msecs) and a mean of 1.2 seconds for the remainder of pauses. For comparison, these short pauses are less than mean written naming response latency for single objects [@torrance2018timed] in a similar population. Longer pauses in the remainder were similar to response latencies for short sentences when describing arrays of images [@roeser2018advance], also in a similar population. Despite the fact that writers can in principle pause when they want to, written composition is often remarkably fluent. The fact that these very short sentence initial pauses exist, and in fact are quite common, is not consistent with the assumption that planning of the upcoming sentence is started and finished in the period between completion of the previous sentence and starting to write the next.

If utterances are not planned in full prior to writing onset, how is it possible that at least for reasonably competent writers producing composition often occurs remarkably fluently, with very few hesitations. The two examples above and similar findings point towards much of the mental activity associated with composition, including the relatively complex processing required to plan sentences, occurring as a result of a cascade of processes that operate partly in parallel and largely without executive control. Consistent with general trends in language processing theory [@dell2007case; @bock2014syntactically; @chang2006], several researchers have argued that the processes associated with written production run at least in part in parallel [@bonin2012evidence; @crump2010hierarchical; @olive2014toward; @van1991handwriting; @roux2013interaction]. @van1991handwriting, for example, argued for a cascade of modular processes, each responsible for a specific transformation (semantic, syntactic, and so forth), with processing occurring as soon as information from the immediately upstream process becomes available [@christiansen2016now]. Buffers provide transient storage at each processing level to accommodate unsynchronised output rates: when lower level processes lag behind higher level processes, buffers allow our fingers to catch up with thought and language related processes. In other words, when writers move from one sentence to another without hesitation this is due to planning the next sentence, to some extent, while completing the output of the previous sentence.

This cascaded account of the composition process gives a rather different understanding of inter-keystroke intervals. IKIs result from one of two underlying mental processes. When upstream processes output at a rate equal to or faster than can be used for finger movement planning, then IKIs are determined just by the time needed for executing finger movements (i.e. just by processing below the last buffer in the processing cascade). However if one or more upstream processes provide output more slowly, then IKIs become dependent directly on upstream processing times and not on finger movement. IKIs therefore form two distributions, one associated with rapid, fluent output, and another that forms as a result of delays caused by some combination of processing at the semantic, syntactic, lexical, or orthographic levels [@roeser2021modelling]. The cascading account provides an explanation for why writers sometimes pause before onsetting a sentence (or word) but often are not. This is not because in some cases sentences are not planned, or planning was postponed until after sentence onset, but that this planning was completed in parallel with previous output.

```{r eval = F}
#Hesitations in writing have potential disadvantages. Writers can, as we have noted, pause at any point without this affecting the eventual communicative effect of their text. However language processing is subject to what @christiansen2016now describe as a fundamental "now-or-never" bottleneck. Buffering is transient and as a consequence of written (and spoken) production are just-in-time systems: Production must flow down the cascade of processes from message to finger movements without significant interruptions. If there is delay -- if, for example, a writer struggles to retrieve the right word or its spelling -- there is a risk that output from upstream processes, and particularly the chain of ideas that the writer wanted to communicate, will be lost. @ronneberg2022process coined this possibility the "process-disruption hypothesis".
```

In language comprehension, Vasishth and colleagues [@vasishth2017feature; @vasishth2017] demonstrated convincingly that some phenomena in online sentence comprehension observed in reading times are captured better by statistical models that assume that the underlying mental process that generates reading data is in fact a mixture of two process expressed as two distributions in the statistical implemental. Similar for keystroke data @roeser2021modelling demonstrated -- using a similar approach as used in the present paper^[For similar approaches see  @almond2012preliminary, @baaijen2012keystroke, @chenu2014interword, @guo2018modeling, @hall2022constructing, @li2021identifying, and @waes2019.] -- how models that assume this mixture of two underlying data generating processes revealed two separate distributions when participants were copy-typing sentences or lists of words that are more pronounced for tasks with challenging orthographic sequences but reduced down to just one distribution in a tapping task where participants merely pressed two keys repeatedly as quickly as possible (e.g. “dkdkdkdk”). As argued by the authors, copy-typing underlies a similar cascade of processes as free text composition. However composition tasks are different from copy tasks: in copy tasks but not in spontaneous text production, content generation and grammatical encoding is artificially constrained to particular sequences of letter bigrams, words or sentences, and copy-typing involves a combination of visual encoding of the target string and mental buffering prior to the activation of motor codes. Typing hesitations in the context of copy-typing may – to some extent – reflect not just production but also, and possibly largely, visual encoding. Text composition, in contrast, provides data from the free generation of words and full sentences that is not dependent on a visually presented text stimulus. For these reasons our research extends the analysis of keystroke data using a mixture model approach to data from spontaneous text production.

In the study that is presented in this paper we test using Bayesian hierarchical models of keystroke intervals -- similar to @roeser2021modelling -- whether the cascading view of writing generalises to contexts in which participants compose texts. On the basis of these statistical models of text composition, we appraise to what extent linguistic edges (sentence and word boundaries) are associated with different pausing behaviours as frequently suggested in the literature [@ailhaud2018variations; @ailhaud2016developmental; @conijn2019understanding; @chukharev2019combined; @torrance2016adolescent; @wen02; @mohsen2021second].

The hypotheses are as follows: if the preparation of the upcoming production unit happens entirely at the corresponding linguistic edge as predicted by the serial account, and not in parallel to production, key-transition intervals can statistically be modelled as a function of the associated transition location. In other words, we expect IKIs to be proportionally longer at before-sentence locations compared to before-word locations compared to within-word locations. This is because transitions before larger linguistic units are associated with processing that involves higher levels of representation. The cascading view, in contrast, assumes that although IKIs at larger linguistic edges might be longer, it is in principle possible that writers do not (always) pause to plan their text but plan in parallel to writing. In other words, statistical models of the pause-and-burst view of written production do not capture that planning can operate, to some extent, in parallel to the output of the written production. Therefore, a statistical model that captures writing as a cascade of processes must account for the possibility that hesitations occur to some extent probabilistically across the entire text although some linguistic locations such as larger linguistic edges are associated with a higher pausing probability. We hypothesise that the statistical model of the parallel cascading view provides a better out-of-sample generalisation than the implementation of the serial model of writing in the context of unconstrained spontaneous text composition.

### 2024-04-28 tutorial/tutorial.Rmd
manuscript update

For example, a mean of 5 msecs on the log scale are `exp(5.85)` $\approx$ `r round(exp(5.85),0)`. This prior with varying degrees of freedom is visualised in Figure \@ref(fig:logprior).

@@

```{r loglik, fig.cap="Log-posterior densities of each data point"}
log_lik(fit_gaus) %>% 
  as_tibble() %>% 
  mutate(sample = 1:n()) %>%
  pivot_longer(-sample, 
               names_to = "obs", 
               values_to = "loglik") %>%
  group_by(obs) %>%
  summarise(M = mean(loglik), var = var(loglik)) %>%
  filter(M > -10) %>% # there are some terrible ones for this model :)
  mutate(obs = as.numeric(str_remove(obs, "V"))) %>%
  ggplot(aes(x = obs, y = M, ymin = M-var, ymax = M+var)) +
  geom_pointrange(fatten = .5) +
  labs(x = "Observations", 
       y = "log-posterior predictive density (with variance)")
```

@@

```{r}
log_lik(fit_gaus) %>% 
  as_tibble() %>% 
  mutate(sample = 1:n()) %>%
  pivot_longer(-sample, 
               names_to = "obs", 
               values_to = "loglik") %>%
  group_by(obs) %>%
  summarise(mean_loglik = mean(exp(loglik)),
            var_loglik = var(loglik)) %>% 
  ungroup() %>%
  summarise(elpd_loo = sum(log(mean_loglik)), # sum of mean log lik
            p_loo = sum(var_loglik), # effective number of parameters
            looic = -2 * (elpd_loo - p_loo)) %>% # Bayesian deviance
  pivot_longer(everything(), names_to = "Estimate") %>% 
  mutate(across(value, round, 1))
```

@@

```{r}
posterior_diffs <- posterior_in_msecs %>%
  pivot_longer(beta:theta, names_to = "parameter") %>%
  group_by(location) %>% 
  mutate(id = row_number()) %>%
  pivot_wider(names_from = location, values_from = value) %>%
  select(-id) %>% # drop id column
  mutate(diff_sent = `before-sentence` - `before-word`,
         diff_word = `before-word` - `within-word`,
         parameter = paste0(parameter, "[diff]")) %>% 
  select(parameter, starts_with("diff"))

@@

# Summarise differences
posterior_diffs %>%
  pivot_longer(starts_with("diff")) %>% 
  group_by(parameter, name) %>%      # group by parameters, comparison
  summarise(across(value, 
            list(mean = base::mean,  # mean difference
                 lower = lower,      # 2.5% lower bound of difference
                 upper = upper,      # 97.5% upper bound of difference
                 p_diff = p_diff),   # prob that difference is negative
            .names = "{.fn}")) %>% 
  ungroup() %>% 
  mutate(across(where(is.numeric), round, 2)) %>% # round after 2nd decimal place
  arrange(name)
```

```{r echo=F}
ps <- posterior_diffs %>%
  pivot_longer(-parameter) %>% 
  group_by(parameter, name) %>%
  summarise(across(value, 
            list(mean = base::mean,  
                 lower = lower,      
                 upper = upper,      
                 p_diff = p_diff),   
            .names = "{.fn}")) %>% 
  ungroup() %>% 
  mutate(across(where(is.numeric), round, 2),
         parameter = gsub("\\[diff\\]", "", parameter)) 
```

@@

```{r}
posterior_ppts_long <- posterior_ppts %>%
  pivot_longer(everything()) %>%
  # separate into location and ppt using "," as separator
  separate(name, into = c("location", "participant"), sep = ",") %>% 
  # extract number from new variables and recode location
  mutate(across(c(location, participant), parse_number),
         across(location, recode, 
                `1` = "before-sentence",
                `2` = "before-word",
                `3` = "within-word")) %>%
  # create different columns for location
  group_by(location) %>% 
  mutate(id = row_number()) %>%
  pivot_wider(names_from = location, values_from = value) %>% 
  select(-id) # drop id

@@

```{r}
posterior_ppts_long_summary <- posterior_ppts_long %>%
  pivot_longer(-participant, names_to = "location") %>%
  group_by(location, participant) %>%
  summarise(across(value, list(mean = base::mean, 
                               lower = lower, 
                               upper = upper), .names = '{.fn}')) %>% 
  ungroup()

@@

```{r}
posterior_ppts_long_diff_summary <- posterior_ppts_long %>% 
  mutate(diff_sent = `before-sentence` - `before-word`,
         diff_word = `before-word` - `within-word`) %>% 
  select(participant, starts_with('diff')) %>%
  pivot_longer(-participant, 
               names_to = 'difference', 
               values_to = 'diff') %>% 
  group_by(participant, difference) %>%
  summarise(across(diff, list(mean = base::mean, 
                               lower = lower, 
                               upper = upper), .names = '{.fn}')) %>% 
  ungroup()

>>>>

For example, a mean of 5 msecs on the log scale are `exp(5.85)` $\approx$ `r round(exp(5.85),0)` msecs on a linear scale. This prior with varying degrees of freedom is visualised in Figure \@ref(fig:logprior).

@@

```{r loglik, fig.cap="Log-posterior densities of each data point"}
log_lik(fit_gaus) %>% 
  as_tibble() %>% 
  mutate(sample = 1:n()) %>%
  pivot_longer(-sample, 
               names_to = "obs", 
               values_to = "loglik") %>%
  summarise(M = mean(loglik), var = var(loglik), .by = obs) %>%
  filter(M > -10) %>% # there are some terrible ones for this model 
  mutate(obs = as.numeric(str_remove(obs, "V"))) %>%
  ggplot(aes(x = obs, y = M, ymin = M - var, ymax = M + var)) +
  geom_pointrange(fatten = .5) +
  labs(x = "Observations", 
       y = "log-posterior predictive density (with variance)")
```

@@

```{r}
log_lik(fit_gaus) %>% 
  as_tibble() %>% 
  mutate(sample = 1:n()) %>%
  pivot_longer(-sample, 
               names_to = "obs", 
               values_to = "loglik") %>%
  summarise(mean_loglik = mean(exp(loglik)),
            var_loglik = var(loglik),
            .by = obs) %>% 
  summarise(elpd_loo = sum(log(mean_loglik)), # sum of mean log lik
            p_loo = sum(var_loglik), # effective number of parameters
            looic = -2 * (elpd_loo - p_loo)) %>% # Bayesian deviance
  pivot_longer(everything(), names_to = "Estimate") %>% 
  mutate(across(value, round, 1))
```

@@

```{r}
posterior_diffs <- posterior_in_msecs %>%
  pivot_longer(beta:theta, names_to = "parameter") %>%
  mutate(id = row_number(), .by = location) %>%
  pivot_wider(names_from = location, values_from = value) %>%
  select(-id) %>% # drop id column
  mutate(diff_sent = `before-sentence` - `before-word`,
         diff_word = `before-word` - `within-word`,
         parameter = paste0(parameter, "[diff]")) %>% 
  select(parameter, starts_with("diff"))

@@

# Summarise differences
posterior_diffs %>%
  pivot_longer(starts_with("diff")) %>% 
  summarise(across(value, 
            list(mean = base::mean,  # mean of difference
                 lower = lower,      # 2.5% lower bound of difference
                 upper = upper,      # 97.5% upper bound of difference
                 p_diff = p_diff),   # prob that difference is negative
            .names = "{.fn}"),
            .by = c(parameter, name)) %>%   # group by parameters, comparison 
  mutate(across(where(is.numeric), round, 2)) %>% # round after 2nd decimal place
  arrange(name)
```

```{r echo=F}
ps <- posterior_diffs %>%
  pivot_longer(-parameter) %>% 
  summarise(across(value, 
            list(mean = base::mean,  
                 lower = lower,      
                 upper = upper,      
                 p_diff = p_diff),   
            .names = "{.fn}"),
            .by = c(parameter, name)) %>% 
  mutate(across(where(is.numeric), ~round(., 2)),
         across(parameter, ~str_remove(., "\\[diff\\]"))) 
```

@@

```{r}
posterior_ppts_long <- posterior_ppts %>%
  pivot_longer(everything()) %>%
  # separate into location and ppt using "," as separator
  separate(name, into = c("location", "participant"), sep = ",") %>% 
  # extract number from new variables and recode location
  mutate(across(c(location, participant), parse_number),
         across(location, recode, 
                `1` = "before-sentence",
                `2` = "before-word",
                `3` = "within-word")) %>%
  # create different columns for location
  mutate(id = row_number(), .by = location) %>%
  pivot_wider(names_from = location, values_from = value) %>% 
  select(-id) # drop id

@@

```{r}
posterior_ppts_long_summary <- posterior_ppts_long %>%
  pivot_longer(-participant, names_to = "location") %>%
  summarise(across(value, list(mean = base::mean, 
                               lower = lower, 
                               upper = upper),
                   .names = '{.fn}'),
            .by = c(location, participant))

@@

```{r}
posterior_ppts_long_diff_summary <- posterior_ppts_long %>% 
  mutate(diff_sent = `before-sentence` - `before-word`,
         diff_word = `before-word` - `within-word`) %>% 
  select(participant, starts_with('diff')) %>%
  pivot_longer(-participant, 
               names_to = 'difference', 
               values_to = 'diff') %>% 
  summarise(across(diff, list(mean = base::mean, 
                               lower = lower, 
                               upper = upper),
                   .names = '{.fn}'),
            .by = c(participant, difference))
